---
title: "User Guide: Interactive Watershed Mapping & Topology Data Structure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{User Guide: Interactive Watershed Mapping & Topology Data Structure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The **`watershed`** R package provides interactive discovery of USGS Hydrologic Unit Code (HUC) subwatershed boundaries, dynamic feature isolation (such as island landmasses via OpenStreetMap), and spatial projection of hexagonal substrate grid networks.

This User Guide explains:
1. Interactive map exploration and regional search (Polygon, Rectangle, and Pin Marker tools).
2. HUC subwatershed granularity auto-scaling and controls.
3. Hexagonal substrate mesh construction, geometric scale selection, and habitat suitability scoring.
4. The structure and usage of the consolidated single-object **Watershed Topology (.rds)** download.

---

## 1. Interactive Map Exploration & Regional Search

The `watershed` package provides two interactive Shiny applications:

- **`watershedApp()`**: Complete interactive explorer combining Leaflet discovery, HUC selection, feature restriction, hexagon grid diameter sliders, and topology downloads.
- **`leafletApp()`**: Standalone Leaflet map explorer focused on point reverse-geocoding and rubberband polygon regional search.

### Discovering Subwatersheds

1. **Rubberband Polygon & Rectangle Tools:** Use the polygon or rectangle draw tool (top left of Leaflet map) to outline an arbitrary regional boundary. Click **Search Watersheds in Region** to query overlapping subwatersheds.
2. **Single-Click Pin Marker Tool:** Click the Marker (pin) tool on the Leaflet draw toolbar to drop a pin marker. Click **Search Watersheds in Region** to query the encompassing USGS HUC subwatershed for that exact point.
3. **Map Selection Toggling:** Click any subwatershed shape directly on the map to toggle inclusion (purple outline) or exclusion (dashed red outline) in-memory without making network API calls.
4. **No Unintentional Tapping Calls:** Tapping the map background does not generate network API calls, preventing unexpected latency while panning or zooming.
5. **Clearing Regions:** Click **Clear Region** or hit the draw tool again to reset the map and start a fresh search.

---

## 2. HUC Watershed Granularity & Auto-Scaling

USGS NHDPlus organizes watersheds into hierarchical Hydrologic Unit Codes (HUC):

| Level | Code Digits | Description | Example Extent |
|-------|-------------|-------------|----------------|
| **HUC12** | 12 digits | Subwatershed (fine detail) | ~10–40 sq km |
| **HUC10** | 10 digits | Watershed (medium detail) | ~100–400 sq km |
| **HUC08** | 8 digits | Sub-basin (coarse detail) | ~1,000–3,000 sq km |
| **HUC06** | 6 digits | Basin | ~10,000+ sq km |
| **HUC04** | 4 digits | Subregion | ~50,000+ sq km |

### Dynamic Auto-Scaling (Code-Annealing)

When drawing a regional polygon across a large geographic area:
- Small drawn regions (e.g. a small county) or point markers return fine-grained **HUC12** subwatersheds.
- Large drawn regions (spanning multiple counties or states) automatically scale starting queries to broader levels (**HUC06** / **HUC04**) based on region area to avoid fetching thousands of tiny HUC12 shapes.
- `watershed` automatically scales HUC granularity using area-based initial query scaling and **code-annealed prefix matching** (`HUC12` $\rightarrow$ `HUC10` $\rightarrow$ `HUC08` $\rightarrow$ `HUC06` $\rightarrow$ `HUC04` $\rightarrow$ `HUC02`).

### Adjusting Granularity in `watershedApp()`

Use the **HUC Level** slider in the sidebar (default: `8`, range: `2` to `12`, step: `2`):
- **Lower levels (2–6):** Selects broad regional sub-basins (`HUC02`, `HUC04`, or `HUC06`).
- **Higher levels (10–12):** Allows more fine-grained subwatersheds (`HUC10` or `HUC12`).

---

## 3. Hexagonal Substrate Grid & Geometric Scale Selection

Once watershed boundaries are selected, `watershed` can optionally project a regular spatial hexagonal grid (`sfc_POLYGON`) across the boundary extent.

- **Optional Hex Overlay:** Unchecking **Include Hexagonal Grid Overlay** skips hex grid construction, returning pure watershed boundaries, subwatersheds, and habitat features with `hex_overlay = NULL`.
- **Geometric Hexagon Scale (10 to 1,000):** Users adjust the **Approximate Number of Hexagons** slider across geometric steps:
  $$\text{Target Hexagons } N \in \{10, 20, 50, 100, 200, 500, 1000\}$$
  (Default: `100` hexagons).
- **Area-Based Diameter Calculation:** `watershed` computes the bounding box area $A$ of the active region and automatically solves for the cell diameter $d$:
  $$d = \sqrt{\frac{A}{0.65 \times N}}$$
  This guarantees that choosing $N = 100$ hexagons will ALWAYS generate ~100 grid cells regardless of whether the target region is a small island or a multi-state watershed.
- **Gray Color & Background Layering:** The hexagonal substrate grid is rendered in subtle slate gray (`#7F8C8D`) and positioned **behind** the HUC watershed boundaries in both Leaflet maps and `ggplot2` autoplots, keeping watershed boundaries bold and prominent.

- **Habitat Suitability Overlay:** When **Overlay Habitat Features & Landmarks** is enabled, `watershed` queries OpenStreetMap for ecological features and scores each hex cell:
  - **Lakes & Ponds:** +2.0 weight
  - **Waterways & Rivers:** +1.5 weight
  - **Bogs & Wetlands:** +1.8 weight
  - **Shaded Forests:** +1.0 weight
  - **Upland / Open:** 1.0 base weight

---

## 4. Consolidated Watershed Topology Download Object

Clicking **Download Watershed Topology (.rds)** in `watershedApp()` exports a single unified S3 object (`habitat_hex_overlay` or `watershed_hex_overlay`) containing all spatial layers and metadata.

### Downloaded Object Structure

```
topology_object (S3 class: habitat_hex_overlay / watershed_hex_overlay)
├── $layer            : sf / sfc polygon boundary (selected region / feature restriction)
├── $individual_hucs  : sf data frame of individual component subwatersheds
├── $hex_overlay      : sfc polygon geometry list of hexagonal grid cells (or NULL if disabled)
├── $hex_habitat_sf   : sf data frame of hex cells with habitat scores & types (or NULL if disabled)
├── $habitat_sf       : sf data frame of extracted OpenStreetMap habitat polygons
├── $landmarks_sf     : sf data frame of point sighting landmarks
├── $huc_id           : character vector of selected HUC ID(s)
├── $feature_name     : character string of applied feature restriction (or NULL)
└── $hex_diameter     : numeric hexagon extent diameter in degrees (or NULL if disabled)
```

### Inspecting Downloaded Files in R

```{r, eval = FALSE}
library(watershed)
library(sf)
library(ggplot2)

# Load the downloaded topology RDS file
topo <- readRDS("isle_royale_topology.rds")

# 1. View overall S3 class and metadata
class(topo)
cat("HUC ID:", topo$huc_id, "\nFeature:", topo$feature_name, "\n")

# 2. Extract boundary polygon
boundary_sf <- topo$layer

# 3. Extract individual subwatersheds
hucs_sf <- topo$individual_hucs

# 4. Extract hexagonal substrate grid with suitability weights (if enabled)
if (!is.null(topo$hex_habitat_sf)) {
  hex_mesh_sf <- topo$hex_habitat_sf
  head(hex_mesh_sf)
}

# 5. Render static ggplot autoplot directly
autoplot(topo)
```
