---
title: "User Guide: Interactive Watershed Mapping & Topology Data Structure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{User Guide: Interactive Watershed Mapping & Topology Data Structure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The **`watershed`** R package provides interactive discovery of USGS Hydrologic Unit Code (HUC) subwatershed boundaries, dynamic feature isolation (such as island landmasses via OpenStreetMap), and spatial projection of hexagonal substrate grid networks.

This User Guide explains:
1. Interactive map exploration and regional search (Polygon, Rectangle, and Pin Marker tools).
2. HUC subwatershed granularity auto-scaling and controls.
3. Hexagonal substrate mesh construction, geometric scale selection, and habitat suitability scoring.
4. The structure and usage of the consolidated single-object **Watershed Topology (.rds)** download.

---

## 1. Interactive Map Exploration & Regional Search

The `watershed` package provides three interactive Shiny applications:

- **`watershedApp()`**: Complete interactive explorer combining Leaflet discovery, HUC selection, feature restriction, NHD stream flowlines, hexagon grid diameter sliders, dynamic legends, and topology downloads.
- **`streamsApp()`**: Dedicated USGS StreamStats point delineation and NHD flowline explorer.
- **`leafletApp()`**: Standalone Leaflet map explorer focused on point reverse-geocoding and rubberband polygon regional search.

### Discovering Subwatersheds

1. **Rubberband Polygon & Rectangle Tools:** Use the polygon or rectangle draw tool (top left of Leaflet map) to outline an arbitrary regional boundary. Click **Search Watersheds in Region** to query overlapping subwatersheds.
2. **Single-Click Pin Marker Tool:** Click the Marker (pin) tool on the Leaflet draw toolbar to drop a pin marker. Click **Search Watersheds in Region** to query the encompassing USGS HUC subwatershed for that exact point.
3. **Map Selection Toggling:** Click any subwatershed shape directly on the map to toggle inclusion (purple outline) or exclusion (dashed red outline) in-memory without making network API calls.
4. **No Unintentional Tapping Calls:** Tapping the map background does not generate network API calls, preventing unexpected latency while panning or zooming.
5. **Clearing Regions:** Click **Clear Region** or hit the draw tool again to reset the map and start a fresh search.

---

## 2. HUC Watershed Granularity & Auto-Scaling

USGS NHDPlus organizes watersheds into hierarchical Hydrologic Unit Codes (HUC):

| Level | Code Digits | Description | Example Extent |
|-------|-------------|-------------|----------------|
| **HUC12** | 12 digits | Subwatershed (fine detail) | ~10–40 sq km |
| **HUC10** | 10 digits | Watershed (medium detail) | ~100–400 sq km |
| **HUC08** | 8 digits | Sub-basin (coarse detail) | ~1,000–3,000 sq km |
| **HUC06** | 6 digits | Basin | ~10,000+ sq km |
| **HUC04** | 4 digits | Subregion | ~50,000+ sq km |

### Dynamic Auto-Scaling (Code-Annealing)

When drawing a regional polygon across a large geographic area:
- Small drawn regions (e.g. a small county) or point markers return fine-grained **HUC12** subwatersheds.
- Large drawn regions (spanning multiple counties or states) automatically scale starting queries to broader levels (**HUC06** / **HUC04**) based on region area to avoid fetching thousands of tiny HUC12 shapes.
- `watershed` automatically scales HUC granularity using area-based initial query scaling and **code-annealed prefix matching** (`HUC12` $\rightarrow$ `HUC10` $\rightarrow$ `HUC08` $\rightarrow$ `HUC06` $\rightarrow$ `HUC04` $\rightarrow$ `HUC02`).

### Adjusting Granularity in `watershedApp()`

Use the **HUC Level** slider in the sidebar (default: `8`, range: `2` to `12`, step: `2`):
- **Lower levels (2–6):** Selects broad regional sub-basins (`HUC02`, `HUC04`, or `HUC06`).
- **Higher levels (10–12):** Allows more fine-grained subwatersheds (`HUC10` or `HUC12`).

---

## 3. Hexagonal Substrate Grid & Geometric Scale Selection

Once watershed boundaries are selected, `watershed` can optionally project a regular spatial hexagonal grid (`sfc_POLYGON`) across the boundary extent.

- **Optional Hex Overlay:** Unchecking **Include Hexagonal Grid Overlay** skips hex grid construction, returning pure watershed boundaries, subwatersheds, and habitat features with `hex_overlay = NULL`.
- **Geometric Hexagon Scale (10 to 1,000):** Users adjust the **Approximate Number of Hexagons** slider across geometric steps:
  $$\text{Target Hexagons } N \in \{10, 20, 50, 100, 200, 500, 1000\}$$
  (Default: `100` hexagons).
- **Area-Based Diameter Calculation:** `watershed` computes the bounding box area $A$ of the active region and automatically solves for the cell diameter $d$:
  $$d = \sqrt{\frac{A}{0.65 \times N}}$$
  This guarantees that choosing $N = 100$ hexagons will ALWAYS generate ~100 grid cells regardless of whether the target region is a small island or a multi-state watershed.
- **Gray Color & Background Layering:** The hexagonal substrate grid is rendered in subtle slate gray (`#7F8C8D`) and positioned **behind** the HUC watershed boundaries in both Leaflet maps and `ggplot2` autoplots, keeping watershed boundaries bold and prominent.

- **Habitat Suitability Overlay:** When **Overlay Habitat Features & Landmarks** is enabled, `watershed` queries OpenStreetMap for ecological features and scores each hex cell:
  - **Lakes & Ponds:** +2.0 weight
  - **Waterways & Rivers:** +1.5 weight
  - **Bogs & Wetlands:** +1.8 weight
  - **Shaded Forests:** +1.0 weight
  - **Upland / Open:** 1.0 base weight

---

## 4. Consolidated Watershed Topology Download Object

Clicking **Download Watershed Topology (.rds)** in `watershedApp()` exports a single unified S3 object (`habitat_hex_overlay` or `watershed_hex_overlay`) containing all spatial layers and metadata.

### Downloaded Object Structure

```
topology_object (S3 class: habitat_hex_overlay / watershed_hex_overlay)
├── $layer            : sf / sfc polygon boundary (selected region / feature restriction)
├── $individual_hucs  : sf data frame of individual component subwatersheds
├── $flowlines        : sf data frame of intersecting NHD stream flowlines (or NULL)
├── $hex_overlay      : sfc polygon geometry list of hexagonal grid cells (or NULL if disabled)
├── $hex_habitat_sf   : sf data frame of hex cells with habitat scores & types (or NULL if disabled)
├── $habitat_sf       : sf data frame of extracted OpenStreetMap habitat polygons
├── $landmarks_sf     : sf data frame of point sighting landmarks
├── $huc_id           : character vector of selected HUC ID(s)
├── $feature_name     : character string of applied feature restriction (or NULL)
└── $hex_diameter     : numeric hexagon extent diameter in degrees (or NULL if disabled)
```

### Inspecting Downloaded Files in R

```{r, eval = FALSE}
library(watershed)
library(sf)
library(ggplot2)

# Load the downloaded topology RDS file
topo <- readRDS("isle_royale_topology.rds")

# 1. View overall S3 class and metadata
class(topo)
cat("HUC ID:", topo$huc_id, "\nFeature:", topo$feature_name, "\n")

# 2. Extract boundary polygon
boundary_sf <- topo$layer

# 3. Extract individual subwatersheds
hucs_sf <- topo$individual_hucs

# 4. Extract hexagonal substrate grid with suitability weights (if enabled)
if (!is.null(topo$hex_habitat_sf)) {
  hex_mesh_sf <- topo$hex_habitat_sf
  head(hex_mesh_sf)
}

# 5. Extract vector stream flowlines (if enabled)
if (!is.null(topo$flowlines)) {
  streams_sf <- topo$flowlines
  head(streams_sf)
}

# 6. Render static ggplot autoplot directly (includes streams, HUCs, and hex grid)
autoplot(topo)
```

---

## 5. Hydrography, Stream Flowlines & Interactive Legends

The `watershed` package provides seamless visual and analytical stream modeling:

### 1. USGS Hydrography, 3DEP Shaded Relief (DEM) & Topographic Base Maps
- `watershedApp()` and `streamsApp()` incorporate official USGS The National Map web map services:
  - **`CartoDB Positron`:** Minimalist high-contrast light base map (default).
  - **`USGS Shaded Relief (DEM)` (`add_usgs_shaded_relief_layer()`):** High-contrast 3D Elevation Program (3DEP) digital elevation hillshade rendering terrain contours, valleys, and ridge lines with exceptional clarity.
  - **`USGS Topo` (`add_usgs_topo_layer()`):** Complete USGS topographic map integrating elevation contours, geographic names, and terrain shading.
  - **`USGS Hydrography (Streams)` (`add_usgs_hydro_layer()`):** Transparent vector overlay of continuous waterways across the United States.
- Toggle between base layers (*CartoDB Positron*, *USGS Shaded Relief (DEM)*, *USGS Topo*, *OpenStreetMap*, *Satellite Imagery*) using the Leaflet layer control in the top-right corner.

### 2. Interactive NHD Stream Flowlines, In-Memory Caching & User-Defined Extents
- When stream flowlines are enabled via the **Stream Flowline Extent** radio control in `watershedApp()`, `watershed` extracts vector stream flowlines via `nhdplusTools::get_nhdplus()` for the active HUC boundaries.
- **Stream Extent Modes:**
  - **`None` (`extent = "none"`, default):** Starts with no stream flowlines loaded for fast initial rendering and minimal network overhead.
  - **`Constrained to HUC(s)` (`extent = "huc"`):** Restricts the flowline network strictly to the selected HUC boundary polygon(s).
  - **`Extended Bounding Box` (`extent = "bbox"`):** Fetches the broader regional river network across the full rectangular bounding box of the watershed region.
  - **`Buffered HUC Region` (`extent = "buffer"`):** Expands the stream retrieval zone outward by a user-defined distance (`1` to `25 km`, default: `5 km`).
- **High-Performance Per-HUC In-Memory Caching & Two-Stage Reactivity:**
  - Flowlines are cached at the individual HUC level (`.watershed_flowline_cache`). When composing multi-HUC regions, only newly added HUCs are fetched from USGS.
  - Removing a sub-watershed or deselecting a HUC drops its flowlines **instantly in memory with 0 network calls**. Deselecting all HUCs triggers immediate layer wipe.
  - Adjusting the **Min Stream Order** slider filters cached flowlines in memory instantly without making additional network requests.
- **Strahler Stream Order Filtering:** Use the **Min Stream Order** slider (1 to 6, default: `4`) to filter flowlines by branching detail:
  - *Order 1–2:* Small headwater creeks and intermittent tributaries.
  - *Order 3–4:* Medium-sized rivers and main stems (default filter captures order 4+ for high-level structure).
  - *Order 5–6+:* Major continental rivers (e.g. Mississippi, Wisconsin, Ohio).
- Line widths are automatically scaled by stream order, and clicking any stream polyline displays its GNIS name, order, and length.

### 3. Dynamic Map Legend
- Check **Display Map Legend** to display a clean, color-coded legend in the bottom-right corner of the Leaflet map:
  - **Selected HUC Boundary:** Purple outline (`#8E44AD`)
  - **Stream Flowlines (NHD):** Solid blue line (`#0055ff`)
  - **Habitat / Landmarks:** Forest green (`#27AE60`)
  - **Hex Mesh Cells:** Slate gray (`#7F8C8D`)
- The legend dynamically adapts whenever layer checkboxes are toggled in the sidebar.

### 4. Standalone USGS StreamStats Explorer (`streamsApp()`)
- Launch `streamsApp()` to interactively click any point on a river or stream:
  - Automatically queries the USGS StreamStats REST API to delineate the upstream catchment basin.
  - Highlights intersecting vector stream flowlines within the selected search radius.
  - Gracefully catches HTTP 422 warnings when clicking outside state boundaries or away from mapped flowline grid cells.

