---
title: "Creation and Setup of watershed R Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creation and Setup of watershed R Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The **`watershed`** R package was extracted from `ewing` into a standalone repository located at [`~/Documents/GitHub/watershed`](../) and published to GitHub at [byandell/watershed](https://github.com/byandell/watershed).

## 1. Package Architecture & Components

- **R Source Code (`R/`)**:
  - [`R/watershedApp.R`](../R/watershedApp.R), [`R/streamsApp.R`](../R/streamsApp.R) & [`R/leafletApp.R`](../R/leafletApp.R) — Interactive Shiny application modules and wrappers.
  - [`R/watershed.R`](../R/watershed.R) & [`R/leaflet.R`](../R/leaflet.R) — USGS HUC subwatershed retrieval, NHD stream flowline extraction, dynamic legends, spatial clipping, and Leaflet overlays.
  - [`R/streams.R`](../R/streams.R) — USGS StreamStats point delineation, NHD vector flowlines, and USGS Hydrography WMS tile layers.
  - [`R/habitat.R`](../R/habitat.R) — OpenStreetMap habitat feature extraction, landmark geocoding, and substrate suitability scoring.
  - [`R/step_controls.R`](../R/step_controls.R) — Geometric scale slider builders (`10, 20, 50, 100, 200, 500, 1000`) and slider choice parsers.
  - [`R/get_site_cache_file.R`](../R/get_site_cache_file.R) — Multi-landscape spatial cache resolver for `"watershed"`.

- **Package Infrastructure & CI/CD**:
  - [`DESCRIPTION`](../DESCRIPTION), [`NAMESPACE`](../NAMESPACE), [`README.md`](../README.md), and [`AGENTS.md`](../AGENTS.md).
  - [`_pkgdown.yml`](../_pkgdown.yml) & [`.github/workflows/pkgdown.yaml`](../.github/workflows/pkgdown.yaml) — Automated pkgdown site generation and `gh-pages` deployment.

- **Data Caches & Quarto Demos**:
  - Pre-cached spatial layers in [`inst/extdata/`](../inst/extdata/).
  - Standalone launchers [`app.R`](../app.R) & [`inst/watershedApp/app.R`](../inst/watershedApp/app.R).
  - Quarto gallery and tutorial vignette in [`demos/`](../demos/) (`demos/index.qmd`, `demos/watershedApp.qmd`).

## 2. Refinements & Quality Assurance

- **R CMD Check Validation**: Corrected S3 method exports (`@exportS3Method ggplot2::autoplot`), suppressed spherical geometry (`s2`) messages and planar warnings, declared global variables, and verified **`0 errors | 0 warnings | 0 notes (Status: OK)`**.
- **Quarto Gallery & Hotlink Fix**: Formatted card containers in [`demos/index.qmd`](../demos/index.qmd) using 4-colon outer fences (`:::: {.grid}`) and added Bootstrap `.stretched-link` styling so clicking anywhere on a demo card navigates to its page.
- **GitHub Integration**: Initialized remote repository `byandell/watershed`, configured issue tracker links, and published documentation to [https://byandell.github.io/watershed/](https://byandell.github.io/watershed/).

## 3. Granularity Controls & Grid Scaling

- **Transparent Polygon & Watershed Styling**:
  - Configured Leaflet and `ggplot2` autoplots ([`R/leaflet.R`](../R/leaflet.R), [`R/leafletApp.R`](../R/leafletApp.R), [`R/watershed.R`](../R/watershed.R), [`R/habitat.R`](../R/habitat.R)) to use transparent fills (`fillOpacity = 0` / `fill = NA`) for drawn regions, subwatershed boundaries, and hexagonal substrate meshes while maintaining distinct border outlines and click interactivity.

- **Zero-Overhead Code-Annealed HUC Scaling**:
  - Enhanced `get_hucs_from_polygon(polygon_sf, max_hucs = 6)` to auto-scale HUC subwatershed granularity dynamically based on the extent of a user-drawn bounding polygon.
  - Inspects 12-digit HUC string prefixes in memory (`HUC12` $\rightarrow$ `HUC10` $\rightarrow$ `HUC08` $\rightarrow$ `HUC06` $\rightarrow$ `HUC04`) after a single HUC12 spatial AOI query, selecting the finest level with $\le \text{max\_hucs}$ regions.

- **Geometric Hexagon Scale & Gray Background Layering**:
  - Replaced degree diameter sliders with an indexed geometric scale slider (`n_hexagons`, values: `10, 20, 50, 100, 200, 500, 1000`, default: `100`) via [`R/step_controls.R`](../R/step_controls.R).
  - Dynamically computes cell diameter $d = \sqrt{A / (0.65 \times N)}$ based on region bounding box area $A$, guaranteeing consistent grid cell density across both small islands and multi-state regional watersheds.
  - Rendered hex grid in subtle slate gray (`#7F8C8D`) and positioned it **behind** HUC boundaries in Leaflet maps and `ggplot2` autoplots.

- **Consolidated Single-Object Download**:
  - Consolidated sidebar download options in [`R/watershedApp.R`](../R/watershedApp.R) into a single **Download Watershed Topology (.rds)** button (`download_watershed`).
  - Downloads a single unified S3 spatial topology object containing all layers (`layer` boundary `sf`, `individual_hucs` `sf`, `hex_overlay` `sfc`, `hex_habitat_sf` suitability `sf`, `habitat_sf` features `sf`, and `landmarks_sf` POI points).

## 4. Interactive Drawing & Geometry Safety Refinements

- **Native Leaflet Draw Toolbar & Single-Feature Auto-Replacement**:
  - Configured `leaflet.extras::addDrawToolbar()` with `singleFeature = TRUE` and styled stroke colors (`#000000`, `weight = 1.5`).
  - Drawing a new region (Polygon, Rectangle, or Pin Marker) automatically replaces and purges previous drawn shapes on the Leaflet map canvas, preventing lingering outlines.
  - Streamlined toolbar choices to 3 robust native options: **Polygon**, **Rectangle**, and **Pin Marker**.

- **Defensive Polygon Geometry Guards**:
  - Added strict geometry type checks in `get_watershed()` ([`R/watershed.R`](../R/watershed.R)) and `render_huc_shapes()` ([`R/leafletApp.R`](../R/leafletApp.R)).
  - When `huc_layer` or `feature_sf` is a point (e.g. from point selection), `get_watershed()` automatically discards non-polygon geometries and fetches the true `MULTIPOLYGON` USGS watershed boundary from NHDPlus, preventing Leaflet `to_ring.default` rendering errors.

- **Point Selection & Compact Marker Styling**:
  - Configured single-click point selection via the Pin Marker draw tool (`markerOptions`).
  - Styled point selections on the Leaflet map with a compact `3.5px` solid black circle marker (`color = "#000000"`, `fillOpacity = 0.9`).

## 5. HUC Level Selection & In-Memory Spatial Aggregation

- **Sidebar HUC Level Control (`2–12`)**:
  - Configured sidebar slider in [`R/watershedApp.R`](../R/watershedApp.R) (`input$huc_level`, `min = 2, max = 12, value = 8, step = 2`) to select target HUC digit levels (`HUC02` through `HUC12`).
  - Formatted HUC layer names and code identifiers with standard two-digit zero padding (`"huc02"`, `"huc04"`, `"huc06"`, `"huc08"`, `"huc10"`, `"huc12"`).

- **0-API-Call In-Memory Spatial Union Aggregation**:
  - Added `aggregate_hucs(hucs_sf, target_level)` in [`R/leaflet.R`](../R/leaflet.R) for zero-overhead in-memory spatial union (`sf::st_union`) of cached base geometries (`raw_fetched_hucs`).
  - Moving the slider from finer to broader levels (e.g. `12` $\rightarrow$ `10` $\rightarrow$ `8` $\rightarrow$ `6` $\rightarrow$ `4` $\rightarrow$ `2`) re-aggregates shapes in real-time with **0 network API calls**.

- **Area-Based Query Scaling & UI Synchronization**:
  - Scaled initial queries based on drawn polygon extent ($km^2$) to avoid REST API timeouts over multi-state regions.
  - Automatically synced the sidebar slider (`shiny::updateSliderInput`) to match the safest base level determined for the region upon searching.
  - Reset sidebar selectize dropdown choices and selected values (`input$huc12_id`) in sync when clicking **Clear Region** or purging drawn shapes.

## 6. Hydrography, Stream Flowlines & Interactive Legends

### User Prompt & Feature Evolution

> **Prompt:** "I want to use streamsApp.R module in watershedApp.R module to add overlay of flowlines and drainage basins. Critique using USGS National Hydrography Tile Service vs Open Street Map. Enable use of legend as in streamsApp. Come up with a plan to improve watershedApp."

### Critique: USGS Hydrography Tile Service vs. OpenStreetMap

- **USGS National Hydrography Dataset (NHD & WMS Tiles)**:
  - *Authority & Continuity:* Federal gold standard for US waterways, maintaining topological continuity, downstream flow direction, and Strahler stream orders (1–10).
  - *WBD Alignment:* Mathematically derived from the same elevation models as USGS HUC subwatersheds and StreamStats catchments.
  - *Transparent Layering:* `USGSHydroCached` provides transparent map tiles that layer cleanly over any base map (Positron, Satellite, OSM).
- **OpenStreetMap (OSM / `osmdata`)**:
  - *Crowdsourced Attributes:* Excellent local detail in populated areas, but lacks hydrologic flow direction, connectivity verification, and stream order hierarchies in rural/remote zones.
  - *Opaque Base Maps:* Standard OSM tiles are fully opaque; extracting vector streams on-the-fly via Overpass API queries introduces rate limits and high network latency.
- **Adopted Strategy:** Use **USGS Hydrography WMS** for continuous base stream visualization, **`nhdplusTools` vector flowlines** for interactive/order-weighted rivers in selected HUCs, and **OSM (`osmdata`)** strictly for non-hydrological landmass boundaries (e.g. island perimeters).

### Implementation Walkthrough & Shiny Module Composition

To maintain clean separation of concerns and achieve **zero code duplication**, stream extraction and legend management are implemented as a composable Shiny module in [`R/streamsApp.R`](../R/streamsApp.R) consumed directly by [`R/watershedApp.R`](../R/watershedApp.R):

1. **Modular UI Component (`streamsInput(id, mode = c("standalone", "sidebar"))`)**:
   - In `mode = "sidebar"`, renders compact sidebar controls (`stream_extent` radio buttons defaulting to `"none"`, `min_stream_order` defaulting to `4`, `show_legend`, and `clear_streams`) for clean embedding inside parent module UI (`watershedInput`).
   - In `mode = "standalone"`, renders full-page explorer controls including region selector dropdown (`rcode`), buffer radius slider, and dedicated map output (`stream_map`).

2. **Modular Server Component (`streamsServer(id, map_proxy_id, watershed_sf, ...)`)**:
   - Operates in embedded mode when `watershed_sf` and `map_proxy_id` are provided by `watershedServer`: implements a two-stage reactive pipeline that queries `get_watershed_flowlines()` for full geometry, performs stream order filtering in memory (0 API calls on slider drag), and synchronizes Leaflet proxy layers and legends.
   - Operates in standalone mode when launched independently: manages map click events, queries StreamStats pour points, and renders catchments.
   - Returns a reactive list of outputs (`$flowlines`, `$show_flowlines`, `$min_stream_order`, `$show_legend`, `$status`) to parent callers.

3. **Leaflet Proxy & Base Map Capabilities ([`R/leaflet.R`](../R/leaflet.R))**:
   - Upgraded `build_base_map()` with multi-provider base groups (*CartoDB Positron*, *USGS Shaded Relief (DEM)*, *USGS Topo*, *OpenStreetMap*, *Satellite Imagery*) and the transparent USGS Hydrography WMS overlay layer.
   - Added `add_usgs_shaded_relief_layer()` and `add_usgs_topo_layer()` to render official USGS 3DEP elevation hillshades and topographic maps.
   - Added `add_leaflet_flowlines()` to render NHD flowlines weighted by stream order (`streamorde`) with clickable popups (stream names, orders, segment lengths).
   - Added `add_watershed_legend()` to render dynamic, color-coded map legends for active layers.

4. **NHD Stream Flowline Extraction, Per-HUC Caching & Autoplot ([`R/watershed.R`](../R/watershed.R))**:
   - Added `get_watershed_flowlines(watershed_obj, min_stream_order = 1, extent = c("huc", "bbox", "buffer"), buffer_km = 5)` with granular per-HUC caching (`.watershed_flowline_cache`) and fast bounding-box remote queries.
   - **Instant Sub-Watershed Removal:** Decomposing or unselecting individual HUCs reuses cached flowlines without making network calls ($0$ API calls, $0$ ms wait).
   - **Fast Envelope Query & Local Intersection:** Queries USGS using spatial bounding boxes, then clips to exact HUC geometries locally in R using `sf::st_intersects` under S2 hygiene.
   - Updated `autoplot.watershed_hex_overlay()` to render vector stream networks underneath the hexagonal mesh in static `ggplot2` plots.

### Empirical Verification Commands

```{r, eval = FALSE}
# 1. R Syntax & Roxygen Verification
parse(file = "R/watershed.R")
parse(file = "R/watershedApp.R")
parse(file = "R/streamsApp.R")
parse(file = "R/leaflet.R")
roxygen2::roxygenise()

# 2. Interactive Map & App Initialization Test
library(leaflet)
devtools::load_all(".")
app1 <- watershedApp()
app2 <- streamsApp()
map <- leaflet() |> add_usgs_hydro_layer() |> add_watershed_legend()
```

